What's in a Name? Linear Temporal Logic Literally Represents Time Lines

### 2023-06-23 paper/main.tex
dfa to nfa

\author{\IEEEauthorblockN{Runming Li}
\IEEEauthorblockA{\textit{Carnegie Mellon University}\\
Pittsburgh, USA \\
runmingl@andrew.cmu.edu}
\and
\IEEEauthorblockN{Keerthana Gurushankar}
\IEEEauthorblockA{\textit{Carnegie Mellon University}\\
Pittsburgh, USA \\
kgurusha@andrew.cmu.edu}
\and
\IEEEauthorblockN{Marijn~J.H.~Heule}
\IEEEauthorblockA{\textit{Carnegie Mellon University}\\
Pittsburgh, USA \\
marijn@cmu.edu}
\and
\IEEEauthorblockN{Kristin~Y.~Rozier}
\IEEEauthorblockA{\textit{Iowa State University} \\
Ames, USA \\
ekyrozier@iastate.edu}
}

\maketitle

@@

\section{Algorithm}
On a high level, our algorithm of converting \ltl\ formulas into timeline visualizations works as follows:
\begin{enumerate}
    \item Convert the given \ltl\ formula to its corresponding \Buchi automaton. % (as detailed in \cref{ltl2aut})
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automaton. % (in \cref{aut2regex})
    \item Simplify the derived $\omega$-regular expression. % (in \cref{regex-simplify})
    \item Visualize the $\omega$-regular expression according to its structure. %(in \cref{regex2timeline})
\end{enumerate}
\begin{figure}[!h]
    \centering
    % \includegraphics[width=\textwidth, scale=0.3]{img/algorithm_outline.pdf}
    TODO
    \caption{Algorithm outline}
    \label{fig:algo}
\end{figure}

@@

\begin{algorithm}[h!]
    \caption{reduce\_dfa}
    \begin{algorithmic}
        \Require ($G, v$), a DFA with state $v$ that is not initial or final
        \Ensure Deletes state $v$ from $G$ while ensuring the same language is recognized
        \For{ every $u\xrightarrow{r_\mathrm{in}} v, v \xrightarrow{r_\mathrm{out}} w$ }
            \If {$v$ has a self-edge  $v\xrightarrow{r_\mathrm{loop}}v$}
                \State replace edges $u\xrightarrow{r_\mathrm{in}} v, v \xrightarrow{r_\mathrm{out}} w$ with $u \xrightarrow{r_\mathrm{in} r_\mathrm{loop}* r_\mathrm{out}} w$
            \Else
                \State replace edges $u\xrightarrow{r_\mathrm{in}} v, v \xrightarrow{r_\mathrm{out}} w$ with $u \xrightarrow{r_\mathrm{in}  r_\mathrm{out}} w$
            \EndIf
        \EndFor
	\State delete node $v$ from $G$
        %\State \Return the reduced graph $G$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[h!]
    \caption{dfa2regex}
    \begin{algorithmic}
        \Require ($G, s, f$), a DFA with initial state $s$ and final state $f$
        \Ensure The regular expression corresponding to all paths from $s$ to $f$
        % WHile there exists any edge except r_sf
        \While{ there exists an interior vertex $v$ }
            \State reduce\_dfa($G, v$)
            \State combine multiedges, i.e. $r1 : u \to w, r2 : u \to w$  to $r1 | r2 : u \to w$
        \EndWhile
        \State \Return $(r_\mathrm{ss}| r_\mathrm{sf} r_\mathrm{ff}^* r_\mathrm{fs})^* r_\mathrm{sf} r_\mathrm{ff}$
        % if called by first visit, this is r_\mathrm{ss}* r_\mathrm{sf}
        % if called by ba2wregex, s=f, this is r_\mathrm{ff}
        % consider modularizing fns differently
        % \State \Return regexp labelling the edge $s\to f$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[h!]
    \caption{dfa2regex\_firstvisit}
    \begin{algorithmic}
        \Require ($G, s, f$), a DFA with initial state $s$ and final state $f$
        \Ensure The regular expression of all paths from $s$ reaching $f$ for the first time
        \State delete all out edges from $f$ in $G$
        \State \Return dfa2regex($G, s, f$)
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[h!]
    \label{alg:ba2wregex}
    \caption{ba2wregex}
    \begin{algorithmic}
        \Require $G$, a \Buchi automaton
        \Ensure The $\omega$-regular expression recognized by $G$
        \State \Return $\bigcup_{f \in F}\left(\text{dfa2regex\_firstvisit}(G, s, f)\text{\^{}dfa2regex}(G, f, f) \right) $
    \end{algorithmic}
\end{algorithm}

@@

\section{Theoretical Analyses} \label{sec:analysis}
\subsection{Correctness of Regex Translation}
\begin{lemma}
    % reduce_dfa does its input to output behaviour
    For any DFA $G$ and any state $v$ in $G$ that is neither a start state or a final state, \textbf{reduce\_dfa}$(G, v)$ preserves the regular language accepted by $G$
\end{lemma}
\begin{proof}
    Let $G'$ be the graph of $G$ post reduction by the application of \textbf{reduce\_dfa}$(G, v)$. We show that the trace of every path accepted by $G$ is also accepted by $G'$. Suppose a path accepted by $G$ does not pass through $v$, clearly this is true. Else, suppose it passes through $v$, since $v$ is an interior node, $v$ cannot be the first or last in the path. Thus, for every pass through $v$, let $u\neq v$ be the last node passed before entering $v$, and likewise $w$ be the first node after exiting $v$. We show that the regular language of sub-traces from $u$ to $w$ in $G$ (shown in \cref{fig:uvw-dfa}) is identical to that in $G'$.
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (u) {$u$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (v) [right=of u] {$v$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (w) [right=of v] {$w$};
	\draw[thick] (u) edge["$r_{uv}$",-latex] (v) (v) edge["$r_{vv}$",-latex,loop above] (v) (v) edge["$r_{vw}$",-latex] (w);
        \end{tikzpicture}
 %       \includegraphics[scale=0.75]{img/uvw_dfa.pdf}
        \caption{A path from $u$ to $w$ in $G$}
        \label{fig:uvw-dfa}
    \end{figure}
    Suppose there is no loop at $v$, the path from $u$ to $w$ must simply be $u\xrightarrow{r_{uv}}v\xrightarrow{r_{vw}}w$, with the trace $r=r_{uv}r_{vw}$. If there is a loop edge $r_{vv}$, this edge can be traveled any number of times before exiting $v$, thus the regular expression is $r=r_{uv}r_{vv}^*r_{vw}$. This edge $u\xrightarrow{r}w$ is added in $G'$. Thus, for every path accepted by $G$, for every subpath entering and exiting $v$ from some $u$ to $w$, there exists a corresponding subpath from $u$ to $w$ in $G'$, and thus a corresponding path accepted by $G'$.
\end{proof}

\begin{lemma}
    For any DFA $G$ with start state $s$ and (exactly one) final state $f$, \textbf{dfa2regex}$(G, s, f)$ outputs the regular expression of all paths from $s$ to $f$ (i.e., the language accepted by $G$)
\end{lemma}
\begin{proof}
    The while loop clearly terminates since the number of interior nodes is strictly decreasing, and upon termination, $G$ is a DFA containing only the nodes $s$ and $f$. Now, every non-terminal visit from $s$ to $f$ and back to $s$, can be replaced by an edge $s\xrightarrow{r_\mathrm{sf}r_\mathrm{ff}^* r_\mathrm{fs}} s$, instead of the edge $f\to s$. This DFA yields the regular language $(r_\mathrm{ss}|r_\mathrm{sf}r_\mathrm{ff}^* r_\mathrm{fs})^* r_\mathrm{sf} r_\mathrm{ff}$, as outputted by the algorithm.
\end{proof}

\begin{lemma}
    For any DFA $G$ with start state $s$ and (exactly one) final state $f$, \textbf{dfa2regex\_firstvisit}$(G, s, f)$ outputs the regular expression of all paths from $s$ to $f$, visiting $f$ for the first time.
\end{lemma}
\begin{proof}
    Let $G'$ be $G$ with all out-edge (including loops) from $f$ deleted. We claim that $L(G')$ is equal to the language $L'$ of all paths from $s$ reaching $f$ for the first time.

@@

\begin{theorem}
    For any \Buchi automaton $G$ with start state $s$ and final states $F$, \textbf{ba2wregex}$(G)$ outputs the $\omega$-regular language accepted by $G$
\end{theorem}
\begin{proof}
    We show that
    \[
        L = \bigcup_{f\in F} \textbf{dfa2regex}(G, s, f) \verb|^| \left(\textbf{dfa2regex\_firstvisit}(G, f, f)\right)^\omega
    \]
    and the $\omega$-language accepted by $G$, $L_\omega(G)$ are identical.

    First, $L\subseteq L_\omega(G)$ since for every infinite run in $L$, clearly by definition of $L$, there is some $f\in F$ such that passes through $f$ infinitely often. The converse is true as well: for every infinite run $q_0, q_1, \dots$ accepted by $G$, then $q_0 = s$ and by definition of the accepting language, there is some accepting state $f$, passed through infinitely often, thus the run is in $\textbf{dfa2regex}(G, s, f)$ \verb|^| $\left(\textbf{dfa2regex\_firstvisit}(G, f, f)\right)^\omega$
\end{proof}
\subsection{Correctness of Rewrite Rules}
\begin{lemma}
    For all regular expressions $r_1, r_2, r$, each of the following rewrite rules preserve the regular or $\omega$-regular language represented by the expressions:
    \begin{alignat}{2}
        & r_1 + r_1r_2^* && \Longrightarrow r_1r_2^* \\
        & r + r && \Longrightarrow r \\
        & r_1 + r_2^*r_1 && \Longrightarrow r_2^*r_1 \\
        & (r^*)^{\omega} && \Longrightarrow r^{\omega} \\
        & (r_1r_2^*)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & (r_1r_2)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & r^*r^{\omega} && \Longrightarrow r^{\omega} \\
        & rr^{\omega} && \Longrightarrow r^{\omega}
    \end{alignat}
\end{lemma}

>>>>

\author{\IEEEauthorblockN{Runming Li\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\textit{Carnegie Mellon University}\\
Pittsburgh, USA \\
runmingl@andrew.cmu.edu}
\and
\IEEEauthorblockN{Keerthana Gurushankar\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\textit{Carnegie Mellon University}\\
Pittsburgh, USA \\
kgurusha@andrew.cmu.edu}
\and 
\IEEEauthorblockN{Marijn~J.H.~Heule}
\IEEEauthorblockA{\textit{Carnegie Mellon University}\\
Pittsburgh, USA \\
marijn@cmu.edu}
\and
\IEEEauthorblockN{Kristin~Y.~Rozier}
\IEEEauthorblockA{\textit{Iowa State University} \\
Ames, USA \\
ekyrozier@iastate.edu}
}

\maketitle
\begingroup\renewcommand\thefootnote{\textsection}
\footnotetext{These authors contributed equally to this work}
\endgroup

@@

\section{Algorithm}
On a high level, our algorithm of converting \ltl\ formulas into timeline visualizations works as follows:
\begin{enumerate}
    \item Convert the given \ltl\ formula to its corresponding \Buchi automaton. % (as detailed in \cref{ltl2aut})
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automaton. % (in \cref{aut2regex})
    \item Simplify the derived $\omega$-regular expression. % (in \cref{regex-simplify})
    \item Visualize the $\omega$-regular expression according to its structure. %(in \cref{regex2timeline})
\end{enumerate}
\begin{figure*}[!t]
    \centering
    \includegraphics[width=\textwidth, scale=0.3]{img/algorithm_outline.pdf}
    \caption{Algorithm outline}
    \label{fig:algo}
\end{figure*}

@@

\begin{algorithm}[h!]
    \caption{reduce\_nfa}
    \begin{algorithmic}
        \Require ($G, v$), an NFA with state $v$ that is not initial or final
        \Ensure Deletes state $v$ from $G$ while ensuring the same language is recognized
        \For{ every $u\xrightarrow{r_\mathrm{in}} v, v \xrightarrow{r_\mathrm{out}} w$ }
            \If {$v$ has a self-edge  $v\xrightarrow{r_\mathrm{loop}}v$}
                \State replace edges $u\xrightarrow{r_\mathrm{in}} v, v \xrightarrow{r_\mathrm{out}} w$ with $u \xrightarrow{r_\mathrm{in} r_\mathrm{loop}* r_\mathrm{out}} w$
            \Else
                \State replace edges $u\xrightarrow{r_\mathrm{in}} v, v \xrightarrow{r_\mathrm{out}} w$ with $u \xrightarrow{r_\mathrm{in}  r_\mathrm{out}} w$
            \EndIf
        \EndFor
	\State delete node $v$ from $G$
        %\State \Return the reduced graph $G$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[h!]
    \caption{nfa2regex}
    \begin{algorithmic}
        \Require ($G, s, f$), an NFA with initial state $s$ and final state $f$
        \Ensure The regular expression corresponding to all paths from $s$ to $f$
        % WHile there exists any edge except r_sf
        \While{ there exists an interior vertex $v$ }
            \State reduce\_nfa($G, v$)
            \State combine multiedges, i.e. $r1 : u \to w, r2 : u \to w$  to $r1 | r2 : u \to w$
        \EndWhile
        \State \Return $(r_\mathrm{ss}| r_\mathrm{sf} r_\mathrm{ff}^* r_\mathrm{fs})^* r_\mathrm{sf} r_\mathrm{ff}$
        % if called by first visit, this is r_\mathrm{ss}* r_\mathrm{sf}
        % if called by ba2wregex, s=f, this is r_\mathrm{ff}
        % consider modularizing fns differently
        % \State \Return regexp labelling the edge $s\to f$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[h!]
    \caption{nfa2regex\_firstvisit}
    \begin{algorithmic}
        \Require ($G, s, f$), an NFA with initial state $s$ and final state $f$
        \Ensure The regular expression of all paths from $s$ reaching $f$ for the first time
        \State delete all out edges from $f$ in $G$
        \State \Return nfa2regex($G, s, f$)
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[h!]
    \label{alg:ba2wregex}
    \caption{ba2wregex}
    \begin{algorithmic}
        \Require $G$, a \Buchi automaton
        \Ensure The $\omega$-regular expression recognized by $G$
        \State \Return $\bigcup_{f \in F}\left(\text{nfa2regex\_firstvisit}(G, s, f)\text{\^{}nfa2regex}(G, f, f) \right) $
    \end{algorithmic}
\end{algorithm}

@@

\section{Theoretical Analyses} \label{sec:analysis}
\subsection{Correctness of Regex Translation}
\begin{lemma}
    % reduce_nfa does its input to output behaviour
    For any NFA $G$ and any state $v$ in $G$ that is neither a start state or a final state, \textbf{reduce\_nfa}$(G, v)$ preserves the regular language accepted by $G$
\end{lemma}
\begin{proof}
    Let $G'$ be the graph of $G$ post reduction by the application of \textbf{reduce\_nfa}$(G, v)$. We show that the trace of every path accepted by $G$ is also accepted by $G'$. Suppose a path accepted by $G$ does not pass through $v$, clearly this is true. Else, suppose it passes through $v$, since $v$ is an interior node, $v$ cannot be the first or last in the path. Thus, for every pass through $v$, let $u\neq v$ be the last node passed before entering $v$, and likewise $w$ be the first node after exiting $v$. We show that the regular language of sub-traces from $u$ to $w$ in $G$ (shown in \cref{fig:uvw-dfa}) is identical to that in $G'$.
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (u) {$u$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (v) [right=of u] {$v$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (w) [right=of v] {$w$};
	\draw[thick] (u) edge["$r_{uv}$",-latex] (v) (v) edge["$r_{vv}$",-latex,loop above] (v) (v) edge["$r_{vw}$",-latex] (w);
        \end{tikzpicture}
 %       \includegraphics[scale=0.75]{img/uvw_dfa.pdf}
        \caption{A path from $u$ to $w$ in $G$}
        \label{fig:uvw-dfa}
    \end{figure}
    Suppose there is no loop at $v$, the path from $u$ to $w$ must simply be $u\xrightarrow{r_{uv}}v\xrightarrow{r_{vw}}w$, with the trace $r=r_{uv}r_{vw}$. If there is a loop edge $r_{vv}$, this edge can be traveled any number of times before exiting $v$, thus the regular expression is $r=r_{uv}r_{vv}^*r_{vw}$. This edge $u\xrightarrow{r}w$ is added in $G'$. Thus, for every path accepted by $G$, for every subpath entering and exiting $v$ from some $u$ to $w$, there exists a corresponding subpath from $u$ to $w$ in $G'$, and thus a corresponding path accepted by $G'$.
\end{proof}

\begin{lemma}
    For any NFA $G$ with start state $s$ and (exactly one) final state $f$, \textbf{nfa2regex}$(G, s, f)$ outputs the regular expression of all paths from $s$ to $f$ (i.e., the language accepted by $G$)
\end{lemma}
\begin{proof}
    The while loop clearly terminates since the number of interior nodes is strictly decreasing, and upon termination, $G$ is an NFA containing only the nodes $s$ and $f$. Now, every non-terminal visit from $s$ to $f$ and back to $s$, can be replaced by an edge $s\xrightarrow{r_\mathrm{sf}r_\mathrm{ff}^* r_\mathrm{fs}} s$, instead of the edge $f\to s$. This NFA yields the regular language $(r_\mathrm{ss}|r_\mathrm{sf}r_\mathrm{ff}^* r_\mathrm{fs})^* r_\mathrm{sf} r_\mathrm{ff}$, as outputted by the algorithm.
\end{proof}

\begin{lemma}
    For any NFA $G$ with start state $s$ and (exactly one) final state $f$, \textbf{nfa2regex\_firstvisit}$(G, s, f)$ outputs the regular expression of all paths from $s$ to $f$, visiting $f$ for the first time.
\end{lemma}
\begin{proof}
    Let $G'$ be $G$ with all out-edge (including loops) from $f$ deleted. We claim that $L(G')$ is equal to the language $L'$ of all paths from $s$ reaching $f$ for the first time.

@@

\begin{theorem}
    For any \Buchi automaton $G$ with start state $s$ and final states $F$, \textbf{ba2wregex}$(G)$ outputs the $\omega$-regular language accepted by $G$
\end{theorem}
\begin{proof}
    We show that
    \[
        L = \bigcup_{f\in F} \textbf{nfa2regex}(G, s, f) \verb|^| \left(\textbf{nfa2regex\_firstvisit}(G, f, f)\right)^\omega
    \]
    and the $\omega$-language accepted by $G$, $L_\omega(G)$ are identical.

    First, $L\subseteq L_\omega(G)$ since for every infinite run in $L$, clearly by definition of $L$, there is some $f\in F$ such that passes through $f$ infinitely often. The converse is true as well: for every infinite run $q_0, q_1, \dots$ accepted by $G$, then $q_0 = s$ and by definition of the accepting language, there is some accepting state $f$, passed through infinitely often, thus the run is in $\textbf{nfa2regex}(G, s, f)$ \verb|^| $\left(\textbf{nfa2regex\_firstvisit}(G, f, f)\right)^\omega$
\end{proof}
\subsection{Correctness of Rewrite Rules}
\begin{lemma}
    For all regular expressions $r_1, r_2, r$, each of the following rewrite rules preserve the regular or $\omega$-regular language represented by the expressions:
    \begin{alignat}{2}
        & r_1 + r_1r_2^* && \Longrightarrow r_1r_2^* \\
        & r + r && \Longrightarrow r \\
        & r_1 + r_2^*r_1 && \Longrightarrow r_2^*r_1 \\
        & (r^*)^{\omega} && \Longrightarrow r^{\omega} \\
        & (r_1r_2^*)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & (r_1r_2)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & r^*r^{\omega} && \Longrightarrow r^{\omega} \\
        & rr^{\omega} && \Longrightarrow r^{\omega}
    \end{alignat}
\end{lemma}
